Equivalence Testing and the Second Generation \emph{P}-Value

### 2018-08-28 manuscript.Rmd

abstract: |
  To move beyond the limitations of null-hypothesis tests, statistical approaches have been developed where the observed data is compared against a range of values that are equivalent to the absence of a meaningful effect. Specifying a range of values around zero allows researchers to statistically reject the presence of effects large enough to matter, and prevents practically insignificant effects from being interpreted as a statistically significant difference. We compare the behavior of the recently proposed second generation *p*-value [@blume_second-generation_2018] with the more established Two One-Sided Tests (TOST) equivalence testing procedure [@schuirmann_comparison_1987]. We show that the two approaches yield almost identical results under optimal conditions. Under suboptimal conditions (e.g., when the confidence interval is wider than the equivalence range, or when confidence intervals are asymmetric) the second generation *p*-value becomes difficult to interpret as a descriptive statistic. The second generation *p*-value is interpretable in a dichotomous manner (i.e., when the SGPV equals 0 or 1 because the confidence intervals lies completely within or outside of the equivalence range), but this dichotomous interpretation does not require calculations. We conclude that equivalence tests yield more consistent *p*-values, distinguish between datasets that yield the same second generation *p*-value, and allow for easier control of Type I and Type II error rates.

@@

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no

@@

```{r, TOSTSGPV6, fig.cap="Comparison of *p*-values from TOST (black line) and SGPV (dotted grey line) across a range of observed sample means (x-axis). Because the sample size is small (n = 10) and the CI is more than twice as wide as the equivalence range (set to -0.4 to 0.4), the SGPV is set to 0.5 (horizontal light grey line) across a range of observed means."}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 600),
     yaxt = "n",
     xaxt = "n",
     ylab = "P-value for TOST and SGPV",
     xlab = "Observed Mean")
axis(1, at = seq(0,600,100), labels = seq(-3,3,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

@@

The effect of asymmetric confidence intervals is most noticeable at smaller sample sizes. In Figure \ref{fig:TOSTSGPV11} we plot the *p*-values from equivalence tests and the SGPV (again plotted as 1-SGPV for ease of comparison) for correlations. The sample size is 30 pairs of observations, and the lower and upper equivalence bounds are set to -0.45 and 0.45, with an alpha of 0.05. As the observed correlation in the sample moves from -1 to 0 the *p*-value from the equivalence test becomes smaller, as does 1-SGPV. The pattern is quite similar to that in Figure \ref{fig:TOSTSGPV2}. The *p*-value for the TOST procedure and 1-SGPV are still identical when *p*-values are 0.975 and 0.025 (indicated by the upper and lower horizontal dotted lines). There are two important differences, however. First of all, the SGPV is no longer a straight line, but a curve, due to the asymmetry in the 95% CI. Second, and most importantly, the *p*-value for the equivalence test and the SGPV do no longer overlap at *p* = 0.5.

>>>>

abstract: |
  To move beyond the limitations of null-hypothesis tests, statistical approaches have been developed where the observed data is compared against a range of values that are equivalent to the absence of a meaningful effect. Specifying a range of values around zero allows researchers to statistically reject the presence of effects large enough to matter, and prevents practically insignificant effects from being interepreted as a statistically significant difference. We compare the behavior of the recently proposed second generation *p*-value [@blume_second-generation_2018] with the more established Two One-Sided Tests (TOST) equivalence testing procedure [@schuirmann_comparison_1987]. We show that the two approaches yield almost identical results under optimal conditions. Under suboptimal conditions (e.g., when the confidence interval is wider than the equivalence range, or when confidence intervals are asymmetric) the second generation *p*-value becomes difficult to interpret as a descriptive statistic. The second generation *p*-value is interpretable in a dichotomous manner (i.e., when the SGPV equals 0 or 1 because the confidence intervals lies completely within or outside of the equivalence range), but this dichotomous interpretation does not require calculations. We conclude that equivalence tests yield more consistent *p*-values, distinguish between datasets that yield the same second generation *p*-value, and allow for easier control of Type I and Type II error rates.

@@

figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
mask              : no

@@

```{r, TOSTSGPV6, fig.cap="Comparison of *p*-values from TOST (black line) and SGPV (dotted grey line) across a range of observed sample means (x-axis). Because the sample size is small (n = 10) and the CI is more than twice as wide as the equivalence range (set to -0.4 to 0.4), the SGPV is set to 0.5 (horizontal lightgrey line) across a range of observed means."}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 600),
     yaxt = "n",
     xaxt = "n",
     ylab = "P-value for TOST and SGPV",
     xlab = "Observed Mean")
axis(1, at = seq(0,600,100), labels = seq(-3,3,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

@@

The effect of asymmetric confidence intervals is most noticable at smaller sample sizes. In Figure \ref{fig:TOSTSGPV11} we plot the *p*-values from equivalence tests and the SGPV (again plotted as 1-SGPV for ease of comparison) for correlations. The sample size is 30 pairs of observations, and the lower and upper equivalence bounds are set to -0.45 and 0.45, with an alpha of 0.05. As the observed correlation in the sample moves from -1 to 0 the *p*-value from the equivalence test becomes smaller, as does 1-SGPV. The pattern is quite similar to that in Figure \ref{fig:TOSTSGPV2}. The *p*-value for the TOST procedure and 1-SGPV are still identical when *p*-values are 0.975 and 0.025 (indicated by the upper and lower horizontal dotted lines). There are two important differences, however. First of all, the SGPV is no longer a straight line, but a curve, due to the asymmetry in the 95% CI. Second, and most importantly, the *p*-value for the equivalence test and the SGPV do no longer overlap at *p* = 0.5.
